|

### 2023-07-15 paper/00-main.Rmd
one minor typo
added new model results to paper

abstract: |
  In American football, a pass rush is an attempt by the defensive team to disrupt the offense and prevent the quarterback (QB) from completing a pass. Existing metrics for assessing pass rush performance are either discrete-time quantities or based on subjective judgement. Using player tracking data, we propose STRAIN, a novel metric for evaluating pass rushers in the National Football League (NFL) at the continuous-time within-play level. Inspired by the concept of strain rate in materials science, STRAIN is a simple and interpretable means for measuring defensive pressure in football. It is a directly-observed statistic as a function of two features: the distance between the pass rusher and QB, and the rate at which this distance is being reduced. Our metric possesses great predictability of pressure and stability over time. We also fit a multilevel model for STRAIN to understand the defensive pressure contribution of every pass rusher at the play-level. We apply our approach to NFL data and present results for the first eight weeks of the 2021 regular season. In particular, we provide comparisons of STRAIN for different defensive positions and play outcomes, and rankings of the NFL's best pass rushers according to our metric.

bibliography: bibliography.bib
output: rticles::asa_article
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.pos = "p"
  #cache = TRUE
)
library(here)
library(tidyverse)
```

\newpage

>>>>

abstract: |
  In American football, a pass rush is an attempt by the defensive team to disrupt the offense and prevent the quarterback (QB) from completing a pass. Existing metrics for assessing pass rush performance are either discrete-time quantities or based on subjective judgment. Using player tracking data, we propose STRAIN, a novel metric for evaluating pass rushers in the National Football League (NFL) at the continuous-time within-play level. Inspired by the concept of strain rate in materials science, STRAIN is a simple and interpretable means for measuring defensive pressure in football. It is a directly-observed statistic as a function of two features: the distance between the pass rusher and QB, and the rate at which this distance is being reduced. Our metric possesses great predictability of pressure and stability over time. We also fit a multilevel model for STRAIN to understand the defensive pressure contribution of every pass rusher at the play-level. We apply our approach to NFL data and present results for the first eight weeks of the 2021 regular season. In particular, we provide comparisons of STRAIN for different defensive positions and play outcomes, and rankings of the NFL's best pass rushers according to our metric.

bibliography: bibliography.bib
biblio-style: chicago
output: rticles::asa_article
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
  # fig.pos = "p",
  # cache = TRUE
)
library(here)
library(tidyverse)
```

\newpage
\spacingset{1.45}

### 2023-07-15 paper/04-results.Rmd
added new model results to paper

We first investigate the predictability of STRAIN as a measure of pressure.
In particular, we look at how well our metric correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided eight-game sample, which reveals a fairly strong correlation ($r =0.625$) between the quantities.
Hence, defenders with high STRAIN values also tend to generate more pressure toward the quarterback.
Furthermore, the average STRAIN for the first four weeks of the 2021 season is more predictive of the last four weeks' pressure rate ($r=0.322$) than the first four weeks' pressure rate ($r=0.097$), as illustrated in Figure \ref{fig:fig_predictability}.

We also analyze the stability of STRAIN over time by comparing the average STRAIN across all frames played for the first and last four weeks of the 2021 NFL regular season.
It is apparent from Figure \ref{fig:fig_stability} that there is a strong positive correlation $(r = 0.854)$ for this relationship.
This means that STRAIN is a highly stable football metric over the provided eight-week time window, and pass rushers appear to carry their STRAIN values with them from week to week.
Overall, STRAIN performs well in both explaining defensive pressure on the field and predicting future performance of pass rushers.

```{r}
#| label: fig_cor_pressure
#| fig.cap: "Relationship between average STRAIN and pressure rate (total hits, sacks, and hurries per snap) over the first eight weeks of the 2021 NFL season. There is a moderately strong association between average STRAIN and pressure rate ($r =0.625$). Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
#| fig.width: 4
#| fig.height: 3
source(here("scripts", "04_fig_cor_pressure.R"))
fig_cor_pressure
```

```{r}
#| label: fig_predictability
#| fig.cap: "Relationships between pressure rate for the last 4 weeks of the 2021 NFL season and first four weeks' pressure rate (left) and average STRAIN (right). STRAIN is more predictive ($r=0.322$) of future pressure rate than previous pressure rate itself ($r=0.097$). Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
#| fig.width: 6.5
#| fig.height: 3
#| fig.show: "hold"
source(here("scripts", "07_fig_predictability.R"))
fig_predictability
```

```{r}
#| label: fig_stability
#| fig.cap: "Relationship between average STRAIN for the last and first four weeks of the 2021 NFL season. A strong linear correlation ($r = 0.854$) demonstrates that STRAIN is a highly stable metric over time. Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
#| fig.width: 4
#| fig.height: 3
source(here("scripts", "05_fig_stability.R"))
fig_stability
```

@@

\begin{table}
\caption{Fixed-effects coefficient estimates for the multilevel model for average STRAIN. Note that the reference down level is first down, the reference pass rusher level is defensive end (DE), denoted R:DE; the reference pass blocker level is tackle (T), denoted B:T. \label{tab:fixedeff}}
\centering
\begin{tabular}{lrrrr}
\hline
& estimate & se & $t$-statistic & $p$-value \\ 
\hline
Intercept & 1.800 & 0.099 & 18.172 & 0.000 \\ 
Number of blockers & -0.737 & 0.026 & -28.415 & 0.000 \\
Yards to go & 0.049 & 0.003 & 14.901 & 0.000 \\
Current yardline & 0.000 & 0.000 & 0.184 & 0.854 \\
$I_{\text{\{2nd down\}}}$ & 0.514 & 0.029 & 17.744 & 0.000 \\
$I_{\text{\{3rd down\}}}$ & 0.918 & 0.032 & 28.361 & 0.000 \\
$I_{\text{\{4th down\}}}$ & 0.465 & 0.077 & 6.064 & 0.000 \\
$I_{\text{\{2pt conversion\}}}$ & -0.138 & 0.180 & -0.770 & 0.441 \\
$I_{\text{\{R:DT\}}}$ & -0.717 & 0.068 & -10.566 & 0.000 \\
$I_{\text{\{R:interior\}}}$ & 0.561 & 0.100 & 5.614 & 0.000 \\
$I_{\text{\{R:NT\}}}$ & -0.943 & 0.102 & -9.234 & 0.000 \\
$I_{\text{\{R:OLB\}}}$ & 0.557 & 0.073 & 7.674 & 0.000 \\
$I_{\text{\{R:secondary\}}}$ & 1.425 & 0.111 & 12.790 & 0.000 \\
$I_{\text{\{B:C\}}}$ & -0.298 & 0.059 & -5.024 & 0.000 \\
$I_{\text{\{B:G\}}}$ & -0.155 & 0.047 & -3.285 & 0.001 \\ 
$I_{\text{\{B:other\}}}$ & -0.126 & 0.065 & -1.951 & 0.051 \\
\hline
\end{tabular}
\end{table}

```{r, eval=FALSE}
                       Estimate Std. Error t value Pr(>|t|)
(Intercept)               1.800      0.099  18.172    0.000
rush_posDT               -0.717      0.068 -10.566    0.000
rush_posinteriorLB        0.561      0.100   5.614    0.000
rush_posNT               -0.943      0.102  -9.234    0.000
rush_posOLB               0.557      0.073   7.674    0.000
rush_possecondary         1.425      0.111  12.790    0.000
block_posC               -0.298      0.059  -5.024    0.000
block_posG               -0.155      0.047  -3.285    0.001
block_posother           -0.126      0.065  -1.951    0.051
n_blockers               -0.737      0.026 -28.415    0.000
down0                    -0.138      0.180  -0.770    0.441
down2                     0.514      0.029  17.744    0.000
down3                     0.918      0.032  28.361    0.000
down4                     0.465      0.077   6.064    0.000
yardsToGo                 0.049      0.003  14.901    0.000
absoluteYardlineNumber    0.000      0.000   0.184    0.854
```

\begin{table}[ht]
\caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}}
\centering
\begin{tabular}{lr}
\hline
Group & ICC \\ 
\hline
Pass rusher & 0.057 \\ 
Pass blocker & 0.012 \\ 
Quarterback & 0.023 \\ 
Residual & 0.907 \\ 
\hline
\end{tabular}
\end{table}

@@

```{r}
#| label: fig_rankings_boot
#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (DE, DT, NT, OLB) by median varying intercept"
#| fig.width: 6
#| fig.height: 5
source(here("scripts", "10_multilevel_mod.R"))
fig_rankings_boot
```

>>>>

We first investigate the predictability of STRAIN as a measure of pressure.
In particular, we look at how well our metric correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided eight-game sample, which reveals a fairly strong correlation ($r =0.6255$) between the quantities.
Hence, defenders with high STRAIN values also tend to generate more pressure toward the quarterback.
Furthermore, the average STRAIN for the first four weeks of the 2021 season is more predictive of the last four weeks' pressure rate ($r=0.3217$) than the first four weeks' pressure rate ($r=0.0965$), as illustrated in Figure \ref{fig:fig_predictability}.

We also analyze the stability of STRAIN over time by comparing the average STRAIN across all frames played for the first and last four weeks of the 2021 NFL regular season.
It is apparent from Figure \ref{fig:fig_stability} that there is a strong positive correlation $(r = 0.8545)$ for this relationship.
This means that STRAIN is a highly stable football metric over the provided eight-week time window, and pass rushers appear to carry their STRAIN values with them from week to week.
Overall, STRAIN performs well in both explaining defensive pressure on the field and predicting future performance of pass rushers.

```{r}
#| label: fig_cor_pressure
#| fig.cap: "Relationship between average STRAIN and pressure rate (total hits, sacks, and hurries per snap) over the first eight weeks of the 2021 NFL season. There is a moderately strong association between average STRAIN and pressure rate ($r =0.6255$). Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
#| fig.width: 4
#| fig.height: 3
source(here("scripts", "04_fig_cor_pressure.R"))
fig_cor_pressure
```

```{r}
#| label: fig_predictability
#| fig.cap: "Relationships between pressure rate for the last 4 weeks of the 2021 NFL season and first four weeks' pressure rate (left) and average STRAIN (right). STRAIN is more predictive ($r=0.3217$) of future pressure rate than previous pressure rate itself ($r=0.0965$). Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
#| fig.width: 6.5
#| fig.height: 3
#| fig.show: "hold"
source(here("scripts", "07_fig_predictability.R"))
fig_predictability
```

```{r}
#| label: fig_stability
#| fig.cap: "Relationship between average STRAIN for the last and first four weeks of the 2021 NFL season. A strong linear correlation ($r = 0.8545$) demonstrates that STRAIN is a highly stable metric over time. Results shown here are for pass rushers with at least 100 snaps played over the eight-week data."
#| fig.width: 4
#| fig.height: 3
source(here("scripts", "05_fig_stability.R"))
fig_stability
```

@@

\begin{table}
\caption{Fixed-effects coefficient estimates for the multilevel model for average STRAIN. Note that the reference down level is first down, the reference pass rusher level is defensive end (DE), denoted R:DE; the reference pass blocker level is tackle (T), denoted B:T. \label{tab:fixedeff}}
\centering
\begin{tabular}{lrrrr}
\hline
& estimate & se & $t$-statistic & $p$-value \\ 
\hline
Intercept & 1.8004 & 0.0991 & 18.1723 & 0.0000 \\ 
Number of blockers & -0.7366 & 0.0259 & -28.4153 & 0.0000 \\
Yards to go & 0.0491 & 0.0033 & 14.9013 & 0.0000 \\
Current yardline & 0.0001 & 0.0005 & 0.1844 & 0.8537 \\
$I_{\text{\{2nd down\}}}$ & 0.5143 & 0.0290 & 17.7439 & 0.0000 \\
$I_{\text{\{3rd down\}}}$ & 0.9176 & 0.0324 & 28.3613 & 0.0000 \\
$I_{\text{\{4th down\}}}$ & 0.4649 & 0.0767 & 6.0637 & 0.0000 \\
$I_{\text{\{2pt conversion\}}}$ & -0.1384 & 0.1797 & -0.7699 & 0.4413 \\
$I_{\text{\{R:DT\}}}$ & -0.7166 & 0.0678 & -10.5658 & 0.0000 \\
$I_{\text{\{R:interior\}}}$ & 0.5611 & 0.0999 & 5.6143 & 0.0000 \\
$I_{\text{\{R:NT\}}}$ & -0.9426 & 0.1021 & -9.2344 & 0.0000 \\
$I_{\text{\{R:OLB\}}}$ & 0.5574 & 0.0726 & 7.6742 & 0.0000 \\
$I_{\text{\{R:secondary\}}}$ & 1.4252 & 0.1114 & 12.7899 & 0.0000 \\
$I_{\text{\{B:C\}}}$ & -0.2983 & 0.0594 & -5.0237 & 0.0000 \\
$I_{\text{\{B:G\}}}$ & -0.1546 & 0.0471 & -3.2847 & 0.0000 \\ 
$I_{\text{\{B:other\}}}$ & -0.1260 & 0.0646 & -1.9508 & 0.0514 \\
\hline
\end{tabular}
\end{table}

<!-- \begin{table}[ht] -->
<!-- \caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}} -->
<!-- \centering -->
<!-- \begin{tabular}{lr} -->
<!-- \hline -->
<!-- Group & ICC \\  -->
<!-- \hline -->
<!-- Pass rusher & 0.057 \\  -->
<!-- Pass blocker & 0.012 \\  -->
<!-- Quarterback & 0.023 \\  -->
<!-- Residual & 0.907 \\  -->
<!-- \hline -->
<!-- \end{tabular} -->
<!-- \end{table} -->

\begin{table}[ht]
\caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}}
\centering
\begin{tabular}{lr}
\hline
Group & ICC \\ 
\hline
Pass rusher & 0.0365 \\ 
Pass blocker & 0.0098 \\ 
Defensive team & 0.0134 \\ 
Offensive team & 0.0143 \\
Residual & 0.9260 \\ 
\hline
\end{tabular}
\end{table}

@@

```{r}
#| label: fig_rankings_boot
#| fig.cap: "Distributions of player effects for top 10 pass rushers in each position (DE, DT, NT, OLB) by median varying intercept."
#| fig.width: 6.3
#| fig.height: 5
source(here("scripts", "10_multilevel_mod.R"))
fig_rankings_boot
```

### 2023-07-15 paper/supplemental.Rmd
added new model results to paper

## Old Multilevel Model for Play-Level STRAIN {#sec:multilevel}

@@

>>>>

# Old Multilevel Model for Play-Level STRAIN {#sec:multilevel}

@@

# Team effects justification

We perform leave-one-week-out cross-validation (LOWO CV)

We evaluate the model performance based on mean squared error (MSE)

```{r}
source(here("scripts", "10_multilevel_mod.R"))
fig_loocv_team_eff
```
